Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-05-30 article/main.tex
fixes

This section describes the \textsf{PCG64} non-cryptographic pseudo-random number
generator (more precisely \textsf{PCG-XSL-RR} in the designer's terminology). It
has an internal state of 128-bit, which operate as a linear congruential
generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]
Where $a = 47026247687942121848144207491837523525$ is a fixed 126-bit
constant. The first initial state $S_0$ is the seed of the generator. The
increment $c$ may be specified by the user of the PRNG to produce different
output streams with the same seed (just as the IV acts in a stream cipher). If
no value of $c$ is specified, then a default increment is provided:
$c = 117397592171526113268558934119004209487$. Note that $c$ must be odd.

@@

In this case, reconstructing the 128-bit internal state $S_i$ of the generator
is sufficient to produce the pseudo-random flow with 100\% accuracy (the
generator can also be clocked backwards if necessary, so that the seed can be
easily reconstructed). We therefore focus on reconstructing $S_0$ (the seed)
from $X_0, X_1, X_2, \dots$. A very simple strategy could be the following:
\begin{enumerate}
\item Guess the 64 upper bits of $S_0$.
\item Compute the missing 64 lower bits using~\eqref{eq:output}, with:
\[
   S_0[0:64] = S_0[64:128] \oplus (X_0  \lll S[122:128]).
\]
\item If $X_1$ is computed correctly, then output $S_0$.
\end{enumerate}

This ``baseline'' procedure requires $2^{64}$ iterations of a loop that does a
dozen arithmetic operations; it always output the correct value of $S_0$, and
may output a few other ones (they can be easily discarded by checking $X_2$). An
improved ``guess-and-determine'' prediction algorithm is possible, which
essentially amounts to expose a truncated version of the underlying linear
congruential generator, and attack it using the well-known tools. This is
possible by combining the following ingredients:
\begin{itemize}
\item The underlying linear congruential generator uses a power-of-two modulus,
  therefore the $\ell$ low-order bits of $S_{i+1}$ are entirely determined by
  the $\ell$ low-order bits of $S_i$. More precisely, we have:
  \begin{equation}\label{eq:lcg}
    S_{i+1} = aS_i + c \bmod 2^\ell, \qquad \text{for all } 0 \leq \ell \leq 128
  \end{equation}
  Therefore, guessing low-order bits of $S_0$ yields a ``long-term advantage''
  that holds for all subsequent states.

@@

Because the borrow are unknown, we in fact compute
$\widetilde{\mathbf{T}'} = \mathbf{S}[58:64+\ell] \boxminus
\mathbf{K}[58:64+\ell]$, and clearly
$\widetilde{\mathbf{T}'} = \mathbf{T}' \boxplus \mathbf{B}$.

@@

    \State $\widetilde T'_j \gets T_j \boxminus K_j[58:64+\ell]$  \Comment{Truncated geometric series}
    \State $(U_0, U_1, U_2) \gets \left\lfloor 2^{58-\ell} \cdot (\widetilde T'_0, \widetilde T'_1, \widetilde T'_2) \cdot  G_3^{-1} \right\rceil \cdot G_3$ \Comment{CVP (Babai rounding)}

@@

% The global strategy to obtain a practical prediction algorithm is essentially
% the same as before (target a truncated version of the underlying linear
% congruential generator), but is a bit more complicated to implement.

Set $\Delta S_i = S_{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
unknown increment $c$). The global ``guess-and-determine'' strategy is
essentially the same as before: gaining access to a truncated version of
$\Delta S_i$, solving a small SVP instance, reconstructing $\Delta S_0$, then
checking consistency.

Let us set:
\begin{align*}
  \nabla S_i \stackrel{def}{=} S_i \boxminus S_0 = \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
\end{align*}
Note that $\nabla S_1 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
\begin{enumerate}
\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{39}$.
\item Reconstruct all rotations $r_i$ from this partial knowledge.
\item Fully reconstruct $\Delta S_0$ from the rotations.
\item Reconstruct $S_0$ from $\Delta S_0$ and the rotations.
\end{enumerate}

\noindent Only the first phase is computationally intensive.

@@

Confirming that the guesses are correct is less immediate. When $c$ was known,
we could reconstruct the internal state; from there, filtering out the bad
guesses was easy. When $c$ is unknown, the same strategy does not work, but a
very strong consistency check can still be implemented. The essential idea is
the following: once $\Delta S_0[0:64+\ell]$ is known, then all the rotations can
be determined accurately and with little computational effort. This in turn
allows a consistency check between the partial information deduced from the
guesses and the actual outputs.

Once we have found $\Delta S_0[0:64+\ell]$, we can compute
$\nabla S_i[0:64+\ell]$ for any $i$ (geometric progression); because we have
guessed the first rotation and the $\ell$ least significant bits of the state,
using~\eqref{eq:output} we gain access to $S_0[58:64+\ell]$; combined with the
``differences'' $\nabla S_i$, this reveals $S_i[58:64+\ell]$ for any $i$ (and we
already had $S_i[0:\ell]$). This allows us to compute
$S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$.

Given a ``fresh'' output $X_i$, and assuming that the guesses are correct, then we should have:
\begin{equation}\label{eq:find_rotation}
  S_i[0:\ell] \oplus S_i[64:64+\ell] = (X_i \lll r_i)[0:\ell].
\end{equation}
If $X_i$ is known, then this gives a way to find the $i$-th rotation $r_i$ by
trying the 64 possible values. As soon as $\ell \geq 6$, this should determine
$r_i$ almost uniquely. In any case, if the guesses were correct, then we should have:
\begin{equation}\label{eq:consistency}
  S_i[0:\ell] \oplus S_i[64:64+\ell] \in \bigl\{ (X_i \lll r)[0:\ell]~|~0 \leq r < 64 \bigr\}.
\end{equation}
If none of the 64 possible rotations yields a match, then the guesses made
beforehand have to be wrong. As a consequence, bad guesses can be filtered
deterministically.

More precisely, let us assume that we have guessed the $\ell$ least-significant
bits of $S_0$ (we denote them by $w_0$) and the first rotation $r_0$. Set
$Y_0 \stackrel{def}{=} X_0 \lll r_0$; there is an unknown value
$x_0$ such that:
\[
  S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
\]
The unknown value $x_0$ will not be recovered until the third phase of the whole
procedure. We obtain the $i$-th state by $S_i \equiv \nabla S_i \boxplus S_0$;
however, because the ``middle'' of $S_0$ is unknown, we have the same problem as
in eq.~\eqref{eq:diff_borrow}: an unknown \emph{carry} may cross the 64-th bit
during the addition and perturb $S_i[64:64+\ell]$. As a result, there is an
unknown vector $\mathbf{C}$, whose components are either 0 or 1, such that such
that:
\begin{align*}
  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (w_0 \oplus Y_0[0:\ell]) \bmod 2^\ell \\
\end{align*}

>>>>

This section describes the \textsf{PCG64} non-cryptographic pseudo-random number
generator (more precisely \textsf{PCG-XSL-RR} in the designer's terminology). It
has an internal state of 128-bit, which operate as a linear congruential
generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]

Where is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
the generator. The increment $c$ may be specified by the user of the PRNG to
produce different output streams with the same seed (just as the IV acts in a
stream cipher). If no value of $c$ is specified, then a default increment is
provided. Note that $c$ must be odd. The default values are:
\begin{align*}
  a &= \phantom{1}47026247687942121848144207491837523525 & \text{(fixed)}\\
  c &= 117397592171526113268558934119004209487           & \text{(default value, user-definable)}
\end{align*}

@@

In this case, reconstructing the 128-bit internal state $S_i$ of the generator
is sufficient to produce the pseudo-random flow with 100\% accuracy (the
generator can also be clocked backwards if necessary, so that the seed can be
easily reconstructed). We therefore focus on reconstructing $S_0$ (the seed)
from $X_0, X_1, X_2, \dots$. A very simple strategy could be the following:
\begin{enumerate}
\item Guess the 64 upper bits of $S_0$ (this includes the rotation).
\item Compute the missing 64 lower bits using~\eqref{eq:output}, with:
\[
   S_0[0:64] = S_0[64:128] \oplus (X_0  \lll S[122:128]).
\]
\item Compute $S_1$ then extract $X_1$; if $X_1$ is correct, then output $S_0$.
\end{enumerate}

This ``baseline'' procedure requires $2^{64}$ iterations of a loop that does a
dozen arithmetic operations; it always output the correct value of $S_0$, and
may output a few other ones (they can be easily discarded by checking $X_2$). An
improved ``guess-and-determine'' prediction algorithm is possible, which
essentially amounts to expose a truncated version of the underlying linear
congruential generator, and attack it using the well-known tools. This is
possible by combining the following ingredients:
\begin{itemize}
\item The underlying linear congruential generator uses a power-of-two modulus,
  therefore the $\ell$ low-order bits of $S_{i+1}$ are entirely determined by
  the $\ell$ low-order bits of $S_i$. More precisely, we have:
  \begin{equation}\label{eq:lcg}
    S_{i+1} = aS_i + c \bmod 2^\ell, \qquad \text{for all } 0 \leq \ell \leq 128
  \end{equation}
  Therefore, guessing the least-significant bits of $S_0$ yields a ``long-term
  advantage'' that holds for all subsequent states.

@@

Because the borrows are unknown, we in fact compute
$\widetilde{\mathbf{T}'} = \mathbf{S}[58:64+\ell] \boxminus
\mathbf{K}[58:64+\ell]$, and clearly
$\widetilde{\mathbf{T}'} = \mathbf{T}' \boxplus \mathbf{B}$.

@@

    \State $\widetilde T'_j \gets T_j \boxminus K_j[58:64+\ell]$  \Comment{Truncated geometric series on $6+\ell$ bits}
    \State $(U_0, U_1, U_2) \gets \left\lfloor 2^{58-\ell} \cdot (\widetilde T'_0, \widetilde T'_1, \widetilde T'_2) \cdot  G_3^{-1} \right\rceil \cdot G_3$ \Comment{CVP (Babai rounding)}

@@

Set $\Delta S_i = S_{i+1} \boxminus S_i$; it is easily checked that $\Delta S_i$ is a
geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
unknown increment $c$). The global ``guess-and-determine'' strategy is
essentially the same as before: gaining access to a truncated version of
$\Delta S_i$, solving a small SVP instance, reconstructing $\Delta S_0$, then
checking consistency.

Let us set:
\begin{equation}\label{eq:nabla}
  \nabla S_i \stackrel{def}{=} S_i - S_0 = \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
\end{equation}
Note that $\nabla S_0 = 0$ and $\nabla S_1 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
\begin{enumerate}
\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{39}$.
\item Reconstruct all rotations $r_i$ from this partial knowledge.
\item Fully reconstruct $\Delta S_0$ from the rotations.
\item Reconstruct $S_0$ from $\Delta S_0$ and the rotations.
\end{enumerate}

\noindent Only the first phase is computationally intensive. The four steps are discussed in the next four subsections.

@@

Confirming that the guesses are correct is less immediate. When $c$ was known,
we could reconstruct the internal state; from there, filtering out the bad
guesses was easy. When $c$ is unknown, the same strategy does not work, but a
very strong consistency check can still be implemented. % The essential idea is
% the following: once $\Delta S_0[0:64+\ell]$ is known, then all the rotations can
% be determined accurately and with little computational effort. This in turn
% allows a consistency check between the partial information deduced from the
% guesses and the actual outputs.

Once we have found $\Delta S_0[0:64+\ell]$, we can compute
$\nabla S_i[0:64+\ell]$ for any $i$ because eq.~\eqref{eq:nabla} holds modulo
$2^{64+\ell}$; because we have guessed the first rotation and the $\ell$ least
significant bits of the state, using~\eqref{eq:output} we gain access to
$S_0[58:64+\ell]$; combined with the ``differences'' $\nabla S_i$, this reveals
$S_i[58:64+\ell]$ for any $i$ (and we already had $S_i[0:\ell]$). This allows us
to compute $S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$. Given a ``fresh''
output $X_i$, and assuming that the guesses are correct, then we should have:
\begin{equation}\label{eq:find_rotation}
  S_i[0:\ell] \oplus S_i[64:64+\ell] = (X_i \lll r_i)[0:\ell].
\end{equation}
% If $X_i$ is known, then this gives a way to find the $i$-th rotation $r_i$ by
% trying the 64 possible values. As soon as $\ell \geq 6$, this should determine
% $r_i$ almost uniquely. 
In particular, if the guesses were correct, then we should have for any $i$:
\begin{equation}\label{eq:consistency}
  S_i[0:\ell] \oplus S_i[64:64+\ell] \in \bigl\{ (X_i \lll r)[0:\ell]~|~0 \leq r < 64 \bigr\}.
\end{equation}
If none of the 64 possible rotations yields a match, then the guesses made
beforehand have to be wrong. As a consequence, bad guesses can be filtered with
an arbitrarily low probability of false positives, by trying several indices
$i$.

A few details still need to be fleshed out. To be precise, let us assume that we
have guessed the $\ell$ least-significant bits of $S_0$ (we denote them by
$w_0$) and the first rotation $r_0$. Set $Y_0 = X_0 \lll r_0$. % there is an
% unknown value $x_0$ such that:
% \[
%   S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
% \]
The unknown value $x_0$ will not be recovered until the third phase of the whole
procedure. We obtain the $i$-th state by $S_i \equiv \nabla S_i \boxplus S_0$;
however, because the ``middle'' of $S_0$ is unknown, we have the same problem as
in eq.~\eqref{eq:diff_borrow}: an unknown \emph{carry} may cross the 64-th bit
during the addition and perturb $S_i[64:64+\ell]$. As a result, there is an
unknown vector $\mathbf{C}$, whose components are either 0 or 1, such that such
that:
\begin{align*}
  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (w_0 \oplus Y_0[0:\ell]) \bmod 2^\ell \\
\end{align*}
